---
title: "Introduction to timecop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to timecop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `timecop` package estimates latent Gaussian vector autoregressive (VAR)
models for discrete-valued multivariate time series. It supports Bernoulli,
Poisson, and Gaussian marginal distributions through a latent variable approach
using Hermite polynomial expansion.

The key idea is that an observed discrete-valued time series
$\{X_t\}$ is generated by an underlying latent Gaussian VAR(1) process
$\{Z_t\}$ via a marginal transformation. The package estimates the VAR(1)
transition matrix of the latent process using a method-of-moments Yule-Walker
procedure, and provides standard errors derived from the long-run variance of
the estimating equations.

## Simulate Data

We begin by simulating bivariate Bernoulli time series data from a known
latent Gaussian VAR(1) process using `latent_var_sim()`:

```{r simulate}
library(timecop)

set.seed(123)
sim <- latent_var_sim(
  d = 2,
  n = 200,
  p = 1,
  param = list(0.5, 0.5),
  phi_lv = matrix(c(0.4, 0.2, 0.2, 0.4), 2, 2),
  family = list("Bernoulli", "Bernoulli")
)

# The simulated observed data is a d x n matrix
dim(sim$X_t)

# First few observations (transposed for readability)
head(t(sim$X_t))
```

## Construct a timecop Object

The `timecop()` constructor takes an n x d data matrix (time points in rows,
variables in columns) and a list of marginal distribution families:

```{r construct}
obj <- timecop(
  data = t(sim$X_t),
  family = list("Bernoulli", "Bernoulli")
)

obj
```

## Fit the Model

Use `fit_timecop()` to estimate the VAR(1) transition matrix and compute
standard errors:

```{r fit}
results <- fit_timecop(obj)
summary(results)

# True values for comparison
cat("\nTrue transition matrix:\n")
print(sim$A_true[,,1])
```

## Supported Distributions

The package currently supports three marginal distributions:

- **Bernoulli**: Binary (0/1) data. The marginal parameter is the success
  probability, estimated as the sample proportion.

- **Poisson**: Count data. The marginal parameter is the rate $\lambda$,
  estimated as the sample mean.

- **Gaussian**: Continuous data. Passed through without transformation (the
  latent and observed processes coincide).

Different variables can have different marginal distributions. For example:

```{r mixed, eval = FALSE}
# Mixed Bernoulli and Poisson
obj_mixed <- timecop(
  data = my_data,
  family = list("Bernoulli", "Poisson", "Poisson")
)
```
